Spatio-angular fluorescence microscopy\\ III. Three-dimensional high-aperture 4f imaging

### 2019-07-29 papers/three-dimensions/paper/paper.tex
Properly normalized monopole tfs and Patrick's comments.

% Talon's custom commands
\setlength\arraycolsep{1.5pt}
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\boldsymbol{\mathfrak{r}}_o}
\providecommand{\rs}{\boldsymbol{\mathfrak{r}}_s}
\newcommand{\mypar}{\parallel}
\providecommand{\ropar}{r_o^{\mypar}}
\providecommand{\roperp}{\mathbf{r}_o^{\bot}}
\providecommand{\rspar}{r_s^{\mypar}}
\providecommand{\rsperp}{\mathbf{r}_s^{\bot}}
\providecommand{\rppar}{r_p^{\mypar}}
\providecommand{\rpperp}{\mathbf{r}_p^{\bot}}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\boldsymbol{\mathfrak{v}}}
\providecommand{\bvperp}{\bs{\nu}^{\bot}}
\providecommand{\bvpar}{\nu^{\parallel}}
\providecommand{\bt}{\bs{\uptau}}
\providecommand{\btperp}{\bs{\tau}^{\bot}}
\providecommand{\btpar}{\tau^{\mypar}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
\providecommand{\eqp}{\stackrel{(p)}{=}}
\providecommand{\propp}{\stackrel{(p)}{\propto}}

@@

\section{Introduction}
\add{Patrick: Key references for comparison are sheppard1994} \cite{sheppard1994} \add{and arnison2002} \cite{arnison2002}.\\

\noindent Skeleton literature review ahead. \\

@@

\noindent \add{I'm having some difficulty getting units to agree in section 2.2.3, but section 3 and the appendix are fairly consistent.}

\subsection{Three-dimensional monopole transfer functions}
\subsubsection{Monopole transfer functions}
Monopole point spread function
\begin{align}
  g(\rd) = \int_{\mbb{R}}d\ropar\int_{\mbb{R}^2}d\roperp\, h(\rd - \roperp, \ropar)f(\ro),
\end{align}
We can rewrite as 
\begin{align}
  G(\bvperp) = \int_{\mbb{R}}d\bvpar\, H(\bv)F(\bv),
\end{align}
where
\begin{align}
  G(\bvperp) = \int_{\mbb{R}^2}d\mb{r}^{\perp}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bvperp),
\end{align}
\begin{align}
  H(\bv) = \int_{\mbb{R}^3}d\bs{\mf{r}}\, h(\bs{\mf{r}})\, \text{exp}(-2\pi i\bs{\mf{r}}\cdot\bv),
\end{align}
and
\begin{align}
  F(\bv) = \int_{\mbb{R}^3}d\bs{\mf{r}}\, f(\bs{\mf{r}})\, \text{exp}(-2\pi i\bs{\mf{r}}\cdot\bv).
\end{align}

@@

\subsubsection{Monopole coherent transfer functions}
\begin{align}
  |c(\rd - \roperp, \ropar)|^2 = h(\rd - \roperp, \ropar)
\end{align}
\begin{align}
  H(\bv) = \int_{\mbb{R}^3}d\bt\,C(\bt)C^*(\bt - \bv)
\end{align}
\begin{align}
  C(\bt) = \int_{\mbb{R}^3}d\bs{\mf{r}}\, c(\bs{\mf{r}})\,\text{exp}(-2\pi i\bs{\mf{r}}\cdot\bt).\label{eq:monoctf}
\end{align}
This implies that the three-dimensional coherent transfer function is a scaled pupil function on a spherical shell.

@@

\subsubsection{Monopole pupil functions}
% First, consider the field created at $\rs$ by a monopole at $\ro$ 
% \begin{align}
%   U_s(\rs, \ro) \propto \frac{\text{exp}[2\pi i\nu_m|\rs - \ro|]}{|\rs - \ro|},
% \end{align}
% where $\nu_m = n_o/\lambda$. For monopoles near the origin we can find the fields on the Gaussian reference sphere of an objectives lens with focal length $f_0 \gg |\ro|$ by applying the approximation $|\rs - \ro| \approx |\rs| - \rs\cdot\ro/|\rs|$ and dropping second-order amplitude and constant phase terms
% \begin{align}
%   U_s(\rs, \ro) \propto \frac{1}{f_0}\text{exp}[-2\pi i\nu_m(\rs\cdot\ro)/f_0].
% \end{align}
% Next, we expand the coordinates into transverse and axial components
% \begin{align}
%   U_s(\rs, \ro) \propto \frac{1}{f_0}\text{exp}[-2\pi i\nu_m(\rsperp\cdot\roperp + \rspar\ropar)/f_0],
% \end{align}
% then rewrite in pupil coordinates using the relationships $\rsperp = \rpperp$ and $\rspar = \sqrt{f_0^2 - |\rppar|^2}$
% \begin{align}
%   U_p(\rpperp, \roperp, \ropar) \propto \frac{1}{f_0}\text{exp}\left[-2\pi i\nu_m\left(\rpperp\cdot\roperp + \rspar\sqrt{f_0^2 - |\roperp|^2}\right)/f_0\right].
% \end{align}
\add{Patrick: This section follows section 2A in the second paper \cite{chandler2019b} fairly closely with modifications for 3D.}

We can write the field in the pupil plane at point $\rpperp$ due to a monopole at $\ro = (\roperp, \ropar)$ as
\begin{align}
  U_p(\rpperp, \roperp, \ropar) \propto p_d(\rpperp, \ropar)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0}\rpperp\cdot\roperp\right],
\end{align}
where $p_d(\rpperp, \ropar)$ is the \textit{defocused monopole pupil function}  
\begin{align}
  p_d(\rpperp, \ropar) = p(\rpperp)\, \text{exp}\left[-2\pi i\ropar\nu_m\sqrt{1 - (|\rpperp|/f_0)^2}\right],
\end{align}
and $p(\rpperp)$ is the usual pupil function which can be used to model angular cutoff, apodization, or aberrations.

Since the second lens is paraxial, we can model the relationship between the field in the pupil plane and the field on the detector with a scaled Fourier transform
\begin{align}
  U_d(\rd', \roperp, \ropar) \propto \int_{\mbb{R}^2}d\rpperp\,p_d(\rpperp, \ropar)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0}\rpperp\cdot\roperp\right]\text{exp}\left[-2\pi i \frac{n_1}{\lambda f_1}\rpperp\cdot\rd'\right].
\end{align}

If we define $P(\btperp, \ropar)$ as the two-dimensional transverse Fourier transform of the defocused pupil function
\begin{align}
  P_d(\btperp, \ropar) = \int_{\mbb{R}^2}d\rpperp\, p_d(\rpperp, \ropar)\,\text{exp}(-2\pi i \rpperp\cdot\btperp),\label{eq:pupilF}
\end{align}
then we can rewrite the field on the detector as
\begin{align}
  U_d(\rd', \roperp, \ropar) \propto P\left(\frac{n_0}{\lambda f_0}\roperp + \frac{n_1}{\lambda f_1}\rd', \ropar\right).
\end{align}
After rewriting in terms of the magnification $m = -\frac{f_1 n_0}{f_0 n_1}$, demagnifying the coordinates with $\rd = \rd'/m$, and demagnifying the irradiance with $h(\rd - \roperp, \ropar) \propto h'(m[\rd - \roperp], \ropar) $, we find the defocused monopole point spread function is related to the Fourier transform of the defocused monopole pupil function by 
\begin{align}
  h(\rd-\roperp, \ropar) \propto \left|P\left(-\frac{n_0}{\lambda f_0}[\rd - \roperp], \ropar\right)\right|^2.
\end{align}
The monopole point spread function is the absolute square of the monopole coherent spread function so
\begin{align}
  c(\rd - \roperp, \ropar) \propto P\left(-\frac{n_0}{\lambda f_0}[\rd - \roperp], \ropar\right). \label{eq:monocsf}
\end{align}
Our final task is to calculate the monopole coherent transfer function by plugging Eq. \eqref{eq:monocsf} into Eq. \eqref{eq:monoctf} which yields
\begin{align}
  C(\bt) = \int_{\mbb{R}}dr^{\mypar}\int_{\mbb{R}^2}d\mb{r}^{\bot}\, P\left(-\frac{n_0}{\lambda f_0}\mb{r}^{\bot}, r^{\mypar}\right)\,\text{exp}(-2\pi i[\mb{r}^{\bot}\cdot\btperp + r^{\mypar}\btpar]).
\end{align}
Using Eq. \eqref{eq:pupilF} to simplify the transverse integral yields
\begin{align}
  C(\bt) \propto p\left(\frac{\lambda f_0}{n_0}\btperp\right) \int_{\mbb{R}}d r^{\mypar} \,\text{exp}\left(-2\pi i r^{\mypar}\sqrt{\nu_m - |\btperp|^2}\right)\,\text{exp}(-2\pi i r^{\mypar}\btpar)
\end{align}
Finally, we use the Fourier shift theorem to evaluate the axial integral to find that 
\begin{align}
  C(\bt) \propto p\left(\frac{\lambda f_0}{n_0}\btperp\right)\,\delta\left(\btpar - \sqrt{\nu_m^2 - |\btperp|^2}\right).\label{eq:monoctf}
\end{align}

@@

\section{Results}
\subsection{Monopole transfer functions}
We can model an aplanatic microscope imaging monopole emitters with the scalar pupil function
\begin{align}
  p(\rpperp) \propto  \tilde{C}\left(\frac{|\rpperp|}{f_0}\right)\Pi\left(\frac{|\rpperp|}{2f_0\sin\alpha}\right). \label{eq:monopupil}
\end{align}

Plugging Eq. \eqref{eq:monopupil} in Eq. \eqref{eq:monoctf} yields
\begin{align}
  C(\bt) \propto \tilde{C}\left(\frac{|\btperp|}{\nu_m}\right)\,\delta\left(\btpar - \sqrt{\nu_m^2 - |\btperp|^2}\right)\,\Pi\left(\frac{|\btperp|}{\nu_c}\right)
\end{align}
where $\nu_c = 2\text{NA}/\lambda$, $\text{NA} = n_0\sin\alpha$, and 
\begin{align}
  \tilde{C}(x) = (1 - x^2)^{-1/4}.
\end{align}

\add{To make the units check we need to add a normalization factor of $\frac{1}{\nu_c}$ to the coherent transfer function. Arguably this factor should have been included in the Eqs. 22 and 23 of the second paper \cite{chandler2019b}, but we played a bit loose with the proportionality symbol. I'm not sure if there's a more elegant way to motivate this factor of $\frac{1}{\nu_c}$.}
\begin{align}
  C(\bt) \propto \frac{1}{\nu_c}\tilde{C}\left(\frac{|\btperp|}{\nu_m}\right)\,\delta\left(\btpar - \sqrt{\nu_m^2 - |\btperp|^2}\right)\,\Pi\left(\frac{|\btperp|}{\nu_c}\right)
\end{align}

In Appendix \ref{sec:monoauto} we calculate the autocorrelation of the coherent transfer function and show that the monopole transfer function is given by
\begin{align}
  H(\bv) = \delta(\bv) + \frac{4\tau_{\circ}^2\bvpar}{\pi\nu_c^2|\bv|\nu_m}\sqrt{p^2 - 1}\,E\left(\beta_m, \frac{p}{\sqrt{p^2 - 1}}\right),
\end{align}
where $E(\beta_m, k)  = \int_0^{\beta_m} d\phi\sqrt{1 - k^2\sin^2\phi}$ is an incomplete elliptic integral of the second kind and
\begin{align}
  p = \frac{|\bvperp|\nu_\circ}{\bvpar|\bv|},\qquad
  \nu_\circ = \sqrt{\nu_m^2 - (|\bv|/2)^2},\qquad
  \beta_m = \cos^{-1}\left(\frac{1}{p}\left[\frac{2\nu_m}{|\bvpar|}\cos\alpha + 1\right]\right).
\end{align}

We can also investigate the three-dimensional monopole transfer functions under the paraxial approximation by expanding the coherent transfer about $|\btperp| = 0$ and dropping second-order terms
\begin{align}
  C(\bt) \propp \frac{1}{\nu_c}\delta\left(\btpar - \nu_m + |\btperp|^2/2\nu_m\right)\,\Pi\left(\frac{|\btperp|}{\nu_c}\right).
\end{align}
In Appendix B we calculate the autocorrelation of the paraxial coherent transfer and show that the paraxial monopole transfer function is given by
\begin{align}
  H(\bv) \eqp \delta(\bv) + \frac{8\nu_m}{\pi\nu_c^2}\text{nez}\left(\frac{|\bvperp|}{\nu_c}, \frac{2\nu_m|\bvpar|}{\nu_c^2}\right),
\end{align}
where
\begin{align}
 \text{nez}(x, z) = \frac{1}{2|x|}\text{Re}\left[\sqrt{1 - \left(\frac{|z|}{|x|} + |x|\right)^2}\right].
\end{align}

@@

\section{Multidimensional curvilinear delta functions}

\add{TODO: Summarize operational rules for tricky curvilinear delta functions. TBH I haven't mastered these rules yet.}
% In this appendix we review operational rules for manipulating multidimensional curvilinear delta functions. For a gentle introduction see Bracewell \cite{bracewell2004} and for details see Gelfand and Shilov \cite{gelfand2016}. Here we attempt to be brief and accessible.

% $\delta(f(x))$

% $\delta(f(\mb{r}))$

% $\delta(\mb{f}(\mb{r}))$

\section{Three-dimensional monopole transfer functions}\label{sec:monoauto}
\add{I have recently reworked this section with units in mind, and as far as I can tell all of the units are consistent. The paraxial results agree with the literature and when they project to 2D we get a result that is consistent with the previous paper.}

\add{I'm still having problems with the high-NA result, though. When I integrate over the axial coordinate the result is not normalized (see Fig. \ref{fig:compare}), and the result does not agree with Sheppard/Barakat \cite{barakat1963, sheppard1994} who predict a suppressed low-frequency response and an enhanced high-frequency response for high-NA objectives compared to the paraxial approximation. I'm hot on the trail of a few issues, but I'd appreciate any flags.}

>>>>

% Talon's custom commands
\setlength\arraycolsep{1.5pt}
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\boldsymbol{\mathfrak{r}}_o}
\providecommand{\rs}{\boldsymbol{\mathfrak{r}}_s}
\newcommand{\mypar}{\parallel}
\providecommand{\ropar}{r_o^{\mypar}}
\providecommand{\roperp}{\mathbf{r}_o^{\bot}}
\providecommand{\rspar}{r_s^{\mypar}}
\providecommand{\rsperp}{\mathbf{r}_s^{\bot}}
\providecommand{\rppar}{r_p^{\mypar}}
\providecommand{\rpperp}{\mathbf{r}_p^{\bot}}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}^{\bot}_d}
\providecommand{\rdp}{\mathbf{r}^{'\bot}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\boldsymbol{\mathfrak{v}}}
\providecommand{\bvperp}{\bs{\nu}^{\bot}}
\providecommand{\bvpar}{\nu^{\parallel}}
\providecommand{\bt}{\bs{\uptau}}
\providecommand{\btperp}{\bs{\tau}^{\bot}}
\providecommand{\btpar}{\tau^{\mypar}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
\providecommand{\eqp}{\stackrel{(p)}{=}}
\providecommand{\propp}{\stackrel{(p)}{\propto}}

@@

\section{Introduction}
\noindent Skeleton literature review ahead. \\

@@

\subsection{Three-dimensional monopole transfer functions}
\subsubsection{Monopole transfer functions}
We begin by modeling a microscope that is imaging a three-dimensional sample of monopole emitters with a two-dimensional detector using an integral transform
\begin{align}
  g(\rd) = \int_{\mbb{R}}d\ropar\int_{\mbb{R}^2}d\roperp\, h(\rd - \roperp, \ropar)f(\ro),\label{eq:opener}
\end{align}
where $\rd$ is a two-dimensional detector position, $\ro = (\roperp, \ropar)$ is a three-dimensional detector coordinate with its transverse/axial decomposition, $g(\rd)$ is the measured irradiance on the detector, $f(\ro)$ is the monopole density, and $h(\rd - \roperp, \ropar)$ is a three-to-two-dimensional point spread function.

We can use the Fourier-convolution theorem to rewrite Eq. \eqref{eq:opener} as
\begin{align}
  G(\bvperp) = \int_{\mbb{R}}d\bvpar\, H(\bv)F(\bv),
\end{align}
where $\bv = (\bvperp, \bvpar)$ is a three-dimensional spatial frequency coordinate with its transverse/axial decomposition, 
\begin{align}
  G(\bvperp) = \int_{\mbb{R}^2}d\rd\, g(\rd)\, \text{exp}(-2\pi i\rd\cdot\bvperp)
\end{align}
is the scaled irradiance spectrum,
\begin{align}
  H(\bv) = \int_{\mbb{R}^3}d\bs{\mf{r}}\, h(\bs{\mf{r}})\, \text{exp}(-2\pi i\bs{\mf{r}}\cdot\bv)
\end{align}
is the monopole transfer function, and
\begin{align}
  F(\bv) = \int_{\mbb{R}^3}d\bs{\mf{r}}\, f(\bs{\mf{r}})\, \text{exp}(-2\pi i\bs{\mf{r}}\cdot\bv)
\end{align}
is the monopole spectrum. Figure \ref{fig:monopole-block} summarizes the relationships between the object and data representations.

@@

\subsubsection{Monopole coherent transfer functions}
The monopole point spread function can always be written as the absolute square of a monopole coherent spread function
\begin{align}
  |c(\rd - \roperp, \ropar)|^2 = h(\rd - \roperp, \ropar).
\end{align}
By applying the autocorrelation theorem we can use the monopole coherent transfer function as a shortcut for calculating the monopole transfer function
\begin{align}
  H(\bv) = \int_{\mbb{R}^3}d\bt\,C(\bt)C^*(\bt - \bv),
\end{align}
where $\bt = (\btperp, \btpar)$ is a three-dimensional transverse spatial frequency coordinate with its transverse/axial decomposition, and $C(\bt)$ is a coherent transfer function defined by
\begin{align}
  C(\bt) = \int_{\mbb{R}^3}d\bs{\mf{r}}\, c(\bs{\mf{r}})\,\text{exp}(-2\pi i\bs{\mf{r}}\cdot\bt).\label{eq:monoctf}
\end{align}
Figure \ref{fig:monopole-transfer-functions} summarizes the relationships between the transfer functions.

@@

\subsubsection{Monopole pupil functions}
In this section we will relate the monopole transfer functions to physical
parameters. First, consider the field created at $\rs$ by a monopole at $\ro$
\begin{align}
  U_s(\rs, \ro) \propto \frac{\text{exp}[2\pi i\nu_m|\rs - \ro|]}{|\rs - \ro|},
\end{align}
where $\nu_m = n_o/\lambda$ is the radius of the Ewald sphere. For monopoles near the origin we can find the fields on the Gaussian reference sphere of an objectives lens with focal length $f_0 \gg |\ro|$ by applying the approximation $|\rs - \ro| \approx |\rs| - \rs\cdot\ro/|\rs|$ and dropping second-order amplitude and constant phase terms
\begin{align}
  U_s(\rs, \ro) \propto \frac{1}{f_0}\text{exp}[-2\pi i\nu_m(\rs\cdot\ro)/f_0].
\end{align}
Next, we expand the coordinates into transverse and axial components
\begin{align}
  U_s(\rs, \roperp, \ropar) \propto \frac{1}{f_0}\text{exp}[-2\pi i\nu_m(\rsperp\cdot\roperp + \rspar\ropar)/f_0],
\end{align}
then rewrite in pupil coordinates using the relationships $\rsperp = \rpperp$ and $\rspar = \sqrt{f_0^2 - |\rppar|^2}$
\begin{align}
  U_p(\rpperp, \roperp, \ropar) \propto \frac{1}{f_0}\text{exp}\left[-2\pi i\nu_m\left(\rpperp\cdot\roperp + \ropar\sqrt{f_0^2 - |\roperp|^2}\right)/f_0\right].
\end{align}
We can rewrite the field in the pupil plane as
\begin{align}
  U_p(\rpperp, \roperp, \ropar) \propto \frac{1}{f_0}p_d(\rpperp, \ropar)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0}\rpperp\cdot\roperp\right],
\end{align}
where $p_d(\rpperp, \ropar)$ is the \textit{defocused monopole pupil function}  
\begin{align}
  p_d(\rpperp, \ropar) = p(\rpperp)\, \text{exp}\left[-2\pi i\nu_m\ropar\sqrt{1 - (|\rpperp|/f_0)^2}\right],
\end{align}
and $p(\rpperp)$ is the usual pupil function that can be used to model the angular cutoff, apodization, aberration, or phase masks.

Since the second lens is paraxial, we can model the relationship between the field in the pupil plane and the field on the detector with a scaled Fourier transform
\begin{align}
  U_d(\rdp, \roperp, \ropar) \propto \frac{n_1}{\lambda f_0f_1}\int_{\mbb{R}^2}d\rpperp\,p_d(\rpperp, \ropar)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0}\rpperp\cdot\roperp\right]\text{exp}\left[-2\pi i \frac{n_1}{\lambda f_1}\rpperp\cdot\rdp\right].
\end{align}

If we define $P(\btperp, \ropar)$ as the two-dimensional transverse Fourier transform of the defocused pupil function
\begin{align}
  P(\btperp, \ropar) = \int_{\mbb{R}^2}d\rpperp\, p_d(\rpperp, \ropar)\,\text{exp}(-2\pi i \rpperp\cdot\btperp),\label{eq:pupilF}
\end{align}
then we can rewrite the field on the detector as
\begin{align}
  U_d(\rdp, \roperp, \ropar) \propto \frac{n_1}{\lambda f_0f_1}P\left(\frac{n_0}{\lambda f_0}\roperp + \frac{n_1}{\lambda f_1}\rdp, \ropar\right).
\end{align}
After rewriting in terms of the magnification $m = -\frac{f_1 n_0}{f_0 n_1}$, demagnifying the coordinates with $\rd = \rdp/m$, and demagnifying the irradiance with $h(\rd - \roperp, \ropar) \propto h'(m[\rd - \roperp], \ropar) $, we find the defocused monopole point spread function is related to the Fourier transform of the defocused monopole pupil function by 
\begin{align}
  h(\rd-\roperp, \ropar) \propto \left|\frac{n_1}{\lambda f_0f_1}P\left(-\frac{n_0}{\lambda f_0}[\rd - \roperp], \ropar\right)\right|^2.
\end{align}
The monopole point spread function is the absolute square of the monopole coherent spread function so
\begin{align}
  c(\rd - \roperp, \ropar) \propto \frac{n_1}{\lambda f_0f_1}P\left(-\frac{n_0}{\lambda f_0}[\rd - \roperp], \ropar\right). \label{eq:monocsf}
\end{align}
Our final task is to calculate the monopole coherent transfer function by plugging Eq. \eqref{eq:monocsf} into Eq. \eqref{eq:monoctf} which yields
\begin{align}
  C(\bt) \propto \frac{n_1}{\lambda f_0f_1}\int_{\mbb{R}}dr^{\mypar}\int_{\mbb{R}^2}d\mb{r}^{\bot}\, P\left(-\frac{n_0}{\lambda f_0}\mb{r}^{\bot}, r^{\mypar}\right)\,\text{exp}(-2\pi i[\mb{r}^{\bot}\cdot\btperp + r^{\mypar}\btpar]).
\end{align}
Using Eq. \eqref{eq:pupilF} to simplify the transverse integral yields
\begin{align}
  C(\bt) \propto \frac{1}{\nu_m}\, p\left(\frac{\lambda f_0}{n_0}\btperp\right) \int_{\mbb{R}}d r^{\mypar} \,\text{exp}\left(-2\pi i r^{\mypar}\sqrt{\nu_m - |\btperp|^2}\right)\,\text{exp}(-2\pi i r^{\mypar}\btpar).
\end{align}
Finally, we use the Fourier shift theorem to evaluate the axial integral to find that 
\begin{align}
  C(\bt) \propto \frac{1}{\nu_m}p\left(\frac{\lambda f_0}{n_0}\btperp\right)\,\delta\left(\btpar - \sqrt{\nu_m^2 - |\btperp|^2}\right).
\end{align}
This shows that the three-dimensional coherent transfer function is a scaled pupil function on a spherical cap.

@@

\section{Results}
\subsection{Monopole transfer functions}
We can model an aplanatic microscope imaging monopole emitters with the scalar pupil function
\begin{align}
  p(\rpperp) \propto  \tilde{C}\left(\frac{|\rpperp|}{f_0}\right)\Pi\left(\frac{|\rpperp|}{2f_0\sin\alpha}\right), \label{eq:monopupil}
\end{align}
where
\begin{align}
  \tilde{C}(x) = (1 - x^2)^{-1/4}.
\end{align}
Plugging Eq. \eqref{eq:monopupil} in Eq. \eqref{eq:monoctf} yields
\begin{align}
  C(\bt) \propto \frac{1}{\nu_m}\tilde{C}\left(\frac{|\btperp|}{\nu_m}\right)\,\delta\left(\btpar - \sqrt{\nu_m^2 - |\btperp|^2}\right)\,\Pi\left(\frac{|\btperp|}{\nu_c}\right),\label{eq:toexpand}
\end{align}
where $\nu_c = 2\text{NA}/\lambda$ and $\text{NA} = n_0\sin\alpha$. Applying an identity from Appendix \ref{sec:delta} (Eq. \eqref{eq:sphereidentity}) allows us to write the coherent transfer function in spherical coordinates as 
\begin{align}
  C(\bt) \propto \frac{1}{\nu_m}\sqrt{\cos\theta_\tau}\,\delta\left(|\bt| - \nu_m\right)\,\Pi\left(\frac{\theta_\tau}{2\alpha}\right).
\end{align}

In Appendix \ref{sec:monoauto} we calculate the autocorrelation of the coherent transfer function and show that the monopole transfer function is given by
\begin{align}
  H(\bv) = 4\pi\sin^2(\alpha/2)\,\delta(|\bv|) + \frac{|\bvpar|}{\nu_m|\bv|}\sqrt{p^2 - 1}\,E\left(\beta_m, \frac{p}{\sqrt{p^2 - 1}}\right)\infty\left(\frac{|\bvperp|}{\nu_m}, \frac{|\bvpar|}{\nu_m}, \alpha\right),
\end{align}
where $E(\beta_m, k)  = \int_0^{\beta_m} d\phi\sqrt{1 - k^2\sin^2\phi}$ is an incomplete elliptic integral of the second kind and
\begin{align}
  p = \frac{2|\bvperp|}{\bvpar|\bv|}\sqrt{\nu_m^2 - (|\bv|/2)^2},\qquad
  \beta_m = \cos^{-1}\left(\frac{1}{p}\left[\frac{2\nu_m}{|\bvpar|}\cos\alpha + 1\right]\right),\nonumber\\ \infty(x, z, \alpha) = \Pi\left(\frac{|z|}{2(\sqrt{1 - (|x| - \sin\alpha)^2} - \cos\alpha)}\right).
\end{align}

We can also investigate the three-dimensional monopole transfer functions under the paraxial approximation by expanding the argument of the delta function in Eq. \eqref{eq:toexpand} about $|\btperp| = 0$ and dropping fourth- and higher-order terms \begin{align}
  C(\bt) \propp \frac{1}{\nu_m}\delta\left(\btpar - \nu_m + |\btperp|^2/2\nu_m\right)\,\Pi\left(\frac{|\btperp|}{\nu_c}\right).
\end{align}
In Appendix B we calculate the autocorrelation of the paraxial coherent transfer and show that the paraxial monopole transfer function is given by
\begin{align}
  H(\bv) \eqp \frac{\pi\nu_c^2}{4\nu_m^2}\delta(|\bv|) + \frac{4}{\nu_m}\text{nez}\left(\frac{|\bvperp|}{\nu_c}, \frac{2\nu_m|\bvpar|}{\nu_c^2}\right),
\end{align}
where
\begin{align}
 \text{nez}(x, z) = \frac{1}{2|x|}\text{Re}\left[\sqrt{1 - \left(\frac{|z|}{|x|} + |x|\right)^2}\right].
\end{align}

Figure 3 compares the high-aperture transfer function to the paraxial transfer function in their normalized forms. For extremely high numerical apertures $\alpha = \pi/2$ the high-aperture transfer function predicts an axial spatial frequency cutoff that is double the prediction of the paraxial model. Additionally, the high-aperture transfer function predicts a spatial frequency support that is a horn torus at $\alpha = \pi/2$ and a lemon-profiled torus for $\alpha < \pi/2$, while the paraxial transfer function predicts a scaled parabolic-lemon-profiled torus for all $\alpha$.

\begin{figure}
  \centering
  \includegraphics[scale=0.85]{../figures/monopole-psf/dpsf.pdf}
  \caption{Comparison of high-aperture and paraxial three-dimensional transfer functions for $\text{NA} = (1.00, 1.25, 1.50)$, $n_o = 1.5$, and $\lambda = 500\,\text{nm}$. Input spatial frequencies are in  m${}^{-1}$ and the output is in m${}^{+1}$. All transfer functions are normalized so that $\lim_{|{\bvperp}|\rightarrow 0} \int_\mbb{R}d\bvpar H(\bvperp, \bvpar) = 1$.}
  \label{fig:compare}
\end{figure}

@@

\section{Multidimensional curvilinear delta functions}\label{sec:delta}
In this appendix we summarize operational rules for manipulating multidimensional curvilinear delta functions. See Bracewell \cite{bracewell2004} for an intuitive approach to two-dimensional delta functions, and see Gelfand and Shilov \cite{gelfand2016} for a more mathematical discussion of multidimensional delta functions.

Our goal is to interpret delta functions of the form $\delta(\mb{g}(\mb{r}))$ where $\mb{g}: \mbb{R}^N \rightarrow \mbb{R}^M$. Starting with a simple case $N = M = 1$ and $g(r) = r$ we can use the sifting property to interpret the delta function
\begin{align}
  \int_{\mbb{R}}dr\,\delta(r)f(r) = f(0).
\end{align}
When $g(r) = ar$ then we can use a change of variables to show that 
\begin{align}
  \int_{\mbb{R}}dr\,\delta(ar)f(r) = \frac{f(0)}{|a|}.
\end{align}
Intuitively, the function $ar$ crosses its zero with an absolute slope $|a|$ so it sifts less of the function $f(r)$ by a factor of $|a|^{-1}$.

When $g(r)$ is a function with multiple zeros then the delta functions sifts (with appropriate weighting) at all of the function's zeros 
\begin{align}
  \int_{\mbb{R}}dr\,\delta(g(r))f(r) = \sum_{g(r_i) = 0}f(r_i)\left|\frac{dg(r_i)}{dr}\right|^{-1}.\label{eq:zero1}
\end{align}

When $N > 1$ and $M = 1$ we can interpret the delta function as sifting over the (potentially infinite) set of zeros of $g(\mb{r})$
\begin{align}
  \int_{\mbb{R}^N}d\mb{r}\,\delta(g(\mb{r}))f(\mb{r}) = \int_{g^{-1}(0)}d\sigma(\mb{r})f(\mb{r})|\nabla g(\mb{r})|^{-1},\label{eq:zero2}
\end{align}
where $d\sigma(\mb{r})$ is an appropriate measure for the solution set, and $|\nabla g(\mb{r})|$ denotes the gradient magnitude evaluated at $\mb{r}$.

When $M > 1$ then $\mb{g}$ is a vector-valued function, and $\delta(\mb{g}(\mb{r})$ can be used to manipulate the product of multiple delta functions. For example, we can rewrite $\delta(g_1(\mb{r}))\delta(g_2(\mb{r}))$ as $\delta(\mb{g}(\mb{r}))$. When $N = M > 1$ 
we interpret the delta function in a familiar way
\begin{align}
  \int_{\mbb{R}^N}d\mb{r}\,\delta(\mb{g}(\mb{r}))f(\mb{r}) = \int_{\mb{g}^{-1}(0)}d\sigma(\mb{r})f(\mb{r})\left|\frac{\partial(g_1,\dots,g_M)}{\partial(r_1,\dots,r_N)}(\mb{r})\right|^{-1},
\end{align}
where $\left|\frac{\partial(g_1,\dots,g_M)}{\partial(r_1,\dots,r_N)}(\mb{r})\right|$ denotes the Jacobian determinant evaluated at $\mb{r}$.
